scRNAseq benchmark in want of a title

### 2019-11-09 bmc_article.tex

Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets and on methods that do not include exogenous control RNA, i.e. spike-ins (see Table \ref{tab:table1} and Supplementary Figure 1 for a description of the datasets). Since graph-based clustering  \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameter and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

We define a pipeline, in R, as a list of functions executed consecutively on the output of the previous one (Figure 0A). Each step can include various parameters, all of which must be given alternative values. The \textit{runPipeline} command will then proceed in a step-wise fashion to compute only the parts of the pipelines that have not already been computed. In addition, each step can output evaluation metrics, which together provide a multi-layered evaluation, where variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of cluster \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and as shown below, might not always capture improvements in earlier steps of the pipeline.

Another motivation for the present framework is that clustering is relatively fragile to variations in the pipeline. We therefore also wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters. We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure 0B).

\includegraphics[scale=0.8]{figure0}

@@

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure \ref{fig:figure1}B). Our method (scDblFinder) achieved the best accuracy while running in a very reasonable time (Figure \ref{fig:figure1}). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure 1D). We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it consistently did (Figure \ref{fig:figure2}).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/figure1_doublets_files/figure-html/figure1-1.png}}
    \caption{\textbf{Identification of doublet cells. A-B:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for the mixology10x3l (\textbf{A}) and mixology10x5cl (\textbf{B}) datasets. DoubletFinder failed due to an error on the mixology10x3cl dataset. \textbf{C:} Running time of the different methods. \textbf{D} Cell called as doublets by scDblFinder tend to be more often misclassified, i.e. assigned a wrong cluster. The corrected percentage of misclassification is the proportion of times a given cell is misclassified across hundreds of clustering pipelines with various parameters, minus the median misclassification rate of the cell's subpopulation.}
    \label{fig:figure1}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/figure2_filtering_files/figure-html/figure2-1.png}}
    \caption{Proportion of cells filtered out and accuracy of the cluster (under two normalizations) for different sets of filters (see methods).}
    \label{fig:figure2}
\end{figure}

On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further. We found that filtering had a much milder impact than doublet removal (Figure \ref{fig:figure3}), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure \ref{fig:figure3}). However, part of the gain is to be ascribed to the removal of cells from subpopulations (especially in the Zheng datasets) that are difficult to distinguish, and for which more stringent clustering did not improve silhouette width (Supplementary Figure 10A). For these reasons, we therefore recommend using a more lenient filter (e.g. lenient or default filter described here) or a cluster-based filter to avoid most of the subpopulation bias while excluding strong outliers.

@@

We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure 13). For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure \ref{fig:figure3}). Regressing out covariates systematically gave poorer performance on all metrics. sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure \ref{fig:figure3}), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure \ref{fig:figure3} and Supplementary Figure 13). Of note, even though sctransform is meant to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets. Interestingly, some clusters tended to be less well distinguished with sctransform than with non-scaled (and in some cases non-normalized) methods, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/figure3_normalization_files/figure-html/figure3-1.png}}
    \caption{\textbf{Evaluation of normalization procedures:} proportion of the variance in the first 5 components explained by the subpopulations, median ARI, and ARI at the true number of clusters.}
    \label{fig:figure3}
\end{figure}

@@

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations (the latter estimated through edgeR). Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 14A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 15B-D). We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 16-17). We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure \ref{fig:figure4}A). A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure4} and Supplementary Figures 16-17). Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.

\begin{figure}
    \centering
    \includegraphics[scale=0.6]{{main_figures/figure4_selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. rankMean and rankMax refer to, respectively, the mean rank and the maximum rank in the following metrics: variance, standardized variance, sctransform, deviance. \texbf{B:} Accuracy of clustering (at the right number of clusters) based on the given selection methods and number of genes selected.}
    \label{fig:figure4}
\end{figure}

@@

The impact of the dimensionality reduction method was far greater than that of normalization or feature selection (e.g. Supplementary Figure 19). Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure \ref{fig:figure5} and Supplementary Figures 19-21). 
In particular, weighing the components had a strong positive impact on silhouette widths and resulting ARI.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/figure5_dr_files/figure-html/unnamed-chunk-2-1.png}}
    \caption{Evaluation of the dimensionality reduction strategies.}
    \label{fig:figure5}
\end{figure}

>>>>

Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets and on methods that do not include exogenous control RNA, i.e. spike-ins (see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets). Since graph-based clustering  \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the Seurat pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameter and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

We define a pipeline, in R, as a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A). Each step can include various parameters, all of which must be given alternative values. The \textit{runPipeline} command will then proceed in a step-wise fashion to compute only the parts of the pipelines that have not already been computed. In addition, each step can output evaluation metrics, which together provide a multi-layered evaluation, where variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of cluster \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and as shown below, might not always capture improvements in earlier steps of the pipeline.

Another motivation for the present framework is that clustering is relatively fragile to variations in the pipeline. We therefore also wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters. We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (\ref{fig:figure2}B).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth,keepaspectratio]{dataset_description}
    \caption{\textbf{Overview of the benchmark datasets used.}}
    \label{fig:figure1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth,keepaspectratio]{figure0}
    \caption{\textbf{Overview of the pipeComp framework and its application to a scRNAseq clustering pipeline.}}
    \label{fig:figure2}
\end{figure}

@@

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure \ref{fig:figure3}B). Our method (scDblFinder) achieved the best accuracy while running in a very reasonable time (Figure \ref{fig:figure3}). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure 1D). We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it consistently did (Figure \ref{fig:figure4}).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/figure1_doublets_files/figure-html/figure1-1.png}}
    \caption{\textbf{Identification of doublet cells. A-B:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for the mixology10x3l (\textbf{A}) and mixology10x5cl (\textbf{B}) datasets. DoubletFinder failed due to an error on the mixology10x3cl dataset. \textbf{C:} Running time of the different methods. \textbf{D} Cell called as doublets by scDblFinder tend to be more often misclassified, i.e. assigned a wrong cluster. The corrected percentage of misclassification is the proportion of times a given cell is misclassified across hundreds of clustering pipelines with various parameters, minus the median misclassification rate of the cell's subpopulation.}
    \label{fig:figure3}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/figure2_filtering_files/figure-html/figure2-1.png}}
    \caption{Proportion of cells filtered out and accuracy of the cluster (under two normalizations) for different sets of filters (see methods).}
    \label{fig:figure4}
\end{figure}

On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further. We found that filtering had a much milder impact than doublet removal (Figure \ref{fig:figure4}), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure \ref{fig:figure4}). However, part of the gain is to be ascribed to the removal of cells from subpopulations (especially in the Zheng datasets) that are difficult to distinguish, and for which more stringent clustering did not improve silhouette width (Supplementary Figure 10A). For these reasons, we therefore recommend using a more lenient filter (e.g. lenient or default filter described here) or a cluster-based filter to avoid most of the subpopulation bias while excluding strong outliers.

@@

We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure 13). For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure \ref{fig:figure5} and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure \ref{fig:figure5}). Regressing out covariates systematically gave poorer performance on all metrics. sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure \ref{fig:figure5}), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure \ref{fig:figure5} and Supplementary Figure 13). Of note, even though sctransform is meant to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets. Interestingly, some clusters tended to be less well distinguished with sctransform than with non-scaled (and in some cases non-normalized) methods, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/figure3_normalization_files/figure-html/figure3-1.png}}
    \caption{\textbf{Evaluation of normalization procedures:} proportion of the variance in the first 5 components explained by the subpopulations, median ARI, and ARI at the true number of clusters.}
    \label{fig:figure5}
\end{figure}

@@

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations (the latter estimated through edgeR). Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 14A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 15B-D). We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 16-17). We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure \ref{fig:figure6}A). A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure6} and Supplementary Figures 16-17). Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.

\begin{figure}
    \centering
    \includegraphics[scale=0.6]{{main_figures/figure4_selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. rankMean and rankMax refer to, respectively, the mean rank and the maximum rank in the following metrics: variance, standardized variance, sctransform, deviance. \texbf{B:} Accuracy of clustering (at the right number of clusters) based on the given selection methods and number of genes selected.}
    \label{fig:figure6}
\end{figure}

@@

The impact of the dimensionality reduction method was far greater than that of normalization or feature selection (e.g. Supplementary Figure 19). Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure \ref{fig:figure7} and Supplementary Figures 19-21). 
In particular, weighing the components had a strong positive impact on silhouette widths and resulting ARI.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/figure5_dr_files/figure-html/unnamed-chunk-2-1.png}}
    \caption{Evaluation of the dimensionality reduction strategies.}
    \label{fig:figure7}
\end{figure}
